Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to Sections 5 and 6.

To answer \rqref{RQ_CompareToolsProjMngmnt} we need to summarize tools that are
visible in the repositories and that are mentioned during the developer
interviews. Software tools are used to support the development, verification,
maintenance, and evolution of software, software processes, and
artifacts~\cite[p.\ 501]{GhezziEtAl2003}. Development tools support the
development of end products, but do not become part of them, unlike dependencies
that remain in the application once it is released \cite[p.\
506]{GhezziEtAl2003}. The tools found for LBM software packages are as follows:

@@

% \begin{table}[h!]
% \begin{center}
% \begin{tabular}{ p{12 cm}}
% %\toprule
% \textbf{Development Tools}\\
% \midrule
% Continuous Integration, Code Editors, Development Environments, Runtime
% Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
% %\midrule
% \textbf{Dependencies}\\
% \midrule
% Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
% %\midrule
% \textbf{Project Management Tools}\\
% \midrule
% Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
% %\bottomrule
% \end{tabular}
% \caption{Observed development tools, dependencies and project management tools} 
% \label{tbl_tools}
% \end{center}
% \end{table}

Once the data on tools is collected, two specific tools that can be compared to
research software norms are version control and continuous integration.

Development tools support the development of end products, but do not become
part of them, unlike dependencies that remain in the application once it is
released \cite[p.\ 506]{GhezziEtAl2003}. Although not shown in
Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
CI. Code editors and compilers were explicitly noted to have been used by
several packages, and were likely used by all of them. One of the packages
(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
Likewise, the use of proprietary code for verifying the correctness of output
was noted by one of the developers (pyLBM). Similar tools were likely used when
developing the other software packages.

For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
software packages use some sort of build automation tools, most commonly Make.
They all use various technical and domain specific libraries. Technical
libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
Open MPI, PyZMQ). Domain specific libraries include scientific computing
libraries (e.g.\ SciPy).

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed.
Table~\ref{tbl_tools} shows the types of project management tools that were
explicitly noted in the artifacts, web-pages, or interviews with the developers.
As with development tools and dependencies, it is possible that other types of
project management tools are used, but they were not visible in the artifacts to
which we had access.  Collaboration tools, most often email and video, are noted
as being used when developing the software projects. Project management software
was not explicitly mentioned, but it is possible that some of the projects use
such software. Many of the projects are located on GitHub, where the developers
use the platform to help manage their projects, especially bug related issues.
Most of the projects appear to use change tracking and version control tools.
Document generation tools are mentioned in 12 of the 24 projects. The tools
Sphinx and Doxygen are explicitly used in this capacity.

The poor adoption of version control tools that Wilson lamented in 2006
\cite{Wilson2006} has greatly improved in the intervening years.  From
Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
(GitHub, GitLab or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
packages use version control.)  The proliferation of version control tools for
LBM matches the increase in the broader research software community.  A little
over 10 years ago version control was estimated to be used in only 50\% of
research software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
survey in 2018 shows 81\% of developers use a version control system
\cite{AlNoamanyAndBorghi2018}.  \cite{Smith2018} has similar results, showing
version control usage for alive projects in mesh generation, geographic
information systems and statistical software for psychiatry increasing from
75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
a four year period ending in 2018.  (For completeness the same study showed a
decrease in version control usage for seismology software over the same time
period, from 41\% down to 36\%).  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
The high usage of version control tools in LBM software matches the trend in
research software in general.

As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
is rarely used in LBM (3 of 24 packages or 12.5\%). This contrasts with the
frequency with which continuous integration is recommended in research software
development guidelines \cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}.  We could not find published data on the
frequency with which continuous integration is used in general research
software.  Our impression is that like LBM software, other research software
packages lag behind the recommended best practice of employing continuous
integration tools.

>>>>

To answer \rqref{RQ_CompareToolsProjMngmnt} we need to summarize tools that are
visible in the repositories and that are mentioned during the developer
interviews. Software tools are used to support the development, verification,
maintenance, and evolution of software, software processes, and
artifacts~\cite[p.\ 501]{GhezziEtAl2003}. Development tools support the
development of end products, but do not become part of them, unlike dependencies
that remain in the application once it is released \cite[p.\
506]{GhezziEtAl2003}. As an example, the tools found for LBM software packages
are as follows:

@@

Once the data on tools is collected, the use of two specific tools should be
compared to research software norms: version control and continuous integration.
The poor adoption of version control tools that Wilson lamented in 2006
\cite{Wilson2006} has greatly improved in the intervening years.  A little over
10 years ago version control was estimated to be used in only 50\% of research
software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
survey in 2018 shows 81\% of developers use a version control system
\cite{AlNoamanyAndBorghi2018}.  \cite{Smith2018} has similar results, showing
version control usage for alive projects in mesh generation, geographic
information systems and statistical software for psychiatry increasing from
75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
a four year period ending in 2018.  (For completeness the same study showed a
decrease in version control usage for seismology software over the same time
period, from 41\% down to 36\%).  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
For LBM packages 67\% of LBM packages use version control (GitHub, GitLab or
CVS). The high usage of version control tools in LBM software matches the trend
in research software in general.

Continuous integration is rarely used in LBM (3 of 24 packages or 12.5\%). This
contrasts with the frequency with which continuous integration is recommended in
research software development guidelines
\cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}. In
this case, it seems likely that the recommendations are ahead of common
practice.
